\section{Discourse analysis}

\begin{frame}
    \frametitle{Discourse analysis}
    \framesubtitle{Choice of Programming Language}

    \begin{itemize}
        \item We considered Elm, Haskell, PureScript and could have considered
            a non-functional programming language like Rust.
            \pause
        \item We chose to implement the editor in Elm.
            \pause
        \item Advantages
            \begin{itemize}
                \item Elm is a pure functional language.
                    \pause
                \item Elm has expressive types such as Algebraic Data Types.
                    \pause
                \item Elm is intended for front-end web development, and as such
                    is good for making graphical user interfaces.
                    \pause
                \item Previous experience using Elm.
            \end{itemize}
        \item Disadvantages
            \begin{itemize}
                \item Elm is very limited with regards to type-classes. I.e. one
                    cannot make types instances of type-classes, and one cannot
                    define new type-classes.
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Discourse analysis}
    \framesubtitle{Testing}

    \begin{itemize}
        \item Making tests during the development of the editor would have been
            helpful.
            \pause
        \item TDD would likely serve this project well.
            \pause
        \item At times we caught some errors that likely would have been caught
            by tests much earlier.
            \pause
        \item Having tests would make us more confident about the correctness of
            the program.
    \end{itemize}
\end{frame}
